\propertitle

### 2019-07-21 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-20T18:44:47+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Our notation and terminology follow \textsc{iso} standards
\citep{iso1993,iso2006,iso2006b} and Jaynes \citey{jaynes1994_r2003} for
\dobs. We often simply say \enquote{belief} for \enquote{\dob}.

@@

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain areas \mynote{be
  more specific}. For many animal species, the neurons that are recorded
within the area are not -- and at present cannot be -- specifically chosen
from among the rest, owing to several limiting factors; for example,
limitations in how precisely electrodes are inserted or neurons are
targeted by viruses. In fact, the set of recorded neurons may even change
during very long recordings or across experimental sessions. We assume that
there's an area, comprising a network of $N$ neurons, for which we equally
believe that any other sample of size $n$ could have been recorded instead
of the sample of $n$ neurons that was actually recorded. This is our
\enquote{full network}.\mynote{how about calling it \enquote{the pool}?}
The normalized total activity of these $N$ neurons at $t$ is $\yA_{t}$. The
relative frequencies of the various activity levels during the recording
were $(\yFF_{\yA}) \defs \yF$. We don't know the values $\yA_{t}$ at each
$t$, or the frequency distribution $\yF$. We only know for certain that
$\yA_{t} \in \set{0,1/N,\dotsc,(N-1)/N,1}$ and that $N\yA_{t} \ge n\ya_{t}$
for obvious reasons. For the time being we assume that we know $N$; in
\sect*** we discuss the consequences of our lack of precise knowledge about
this number.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those expectations
to be equal to measured averages. We add two remarks about this method that
are seldom made in the literature. First, the distribution $\px(\yA)$ given
by this method is the zeroth-order approximation \citep[in the sense of
Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four completely different quantities. Note that the four
distributions above numerically differ in higher-order approximations. We
discuss their differences in \sect***. Second, the maximum-entropy method
based on the Shannon entropy implicitly makes some assumptions about the
probabilities for the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

In our case, to apply the method for the total activity of the full network
we need to fix some of the latter's averages, for example its moments. But
we don't have any measured moments for the full network to equate the
distribution moments to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact, linear relation between the first $m$
moments for the full network and the first $m$ for the sample
\citep[\eqns~(16)]{portamanaetal2015}; the ones determine the others and
vice versa at every time bin. This relation is a classical result of
sampling theory
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
linearly determine the first $m$ power moments and vice versa, because
$\tbinom{n\ya}{m}$ is a polynomial in $\ya$ of degree $m$; so fixing the
ones is equivalent to fixing the others. Now we have this extremely
convenient property: \emph{the first $n$ normalized factorial moments for a
  sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is the distribution for the full-network activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the maximum-entropy distribution
$\px(\yA)$ for the full network are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)
\end{equation}
for all $m$ we wish.

The calculation amounts to a convex optimization
\citep{meadetal1984}[\chap~10]{pressetal1988_r2007}{fangetal1997,boydetal2004_r2009,portamana2017b}
and for the numbers $N,n,m$ considered in the next section it can be done
on a modern computer without approximations of normalization constants or
potential functions.

@@

The number of moments used with this method depends on the questions and
hypotheses that a researcher is exploring; for example, hypotheses of
sufficient statistics, such as the sufficiency of pairwise correlations to
quantify our \dob\ about network activity. In the present work we
only want to introduce the general method without entering into biological
questions of this kind.

The particular case in which $n$ moments are constrained is especially
important: it corresponds to fully constraining the marginal frequency
distribution for the activity of the sample neurons. In this case, our
belief about the full-network activity is based on all available
measured frequency data. Note that application of the maximum-entropy
method \emph{at the sample level} is trivial and meaningless in this case
-- it just gives back the measured frequency distribution. But application
of the method \emph{at the level of the full network} is not trivial.

@@

In the next section we apply the method just described to the data sets
from two actual recordings, using *** moments, and discuss the properties
of the resulting distributions.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-21T17:05:52+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Our notation and terminology follow \textsc{iso}
\citey{iso1993,iso2006,iso2006b} standards and Jaynes
\citey{jaynes1994_r2003} for \dobs. We often simply say \enquote{belief}
for \enquote{\dob}.

@@

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain areas \mynote{be
  more specific}. For many animal species, the neurons that are recorded
within the area are not -- and at present cannot be -- specifically chosen
from among the rest, owing to several limiting factors; for example,
limitations in how precisely electrodes are inserted or neurons are
targeted by viruses. In fact, the set of recorded neurons may even change
during very long recordings or across experimental sessions. We assume that
there's an area, comprising a network of $N$ neurons, for which we believe
that any other sample of size $n$ could have equally plausibly been
recorded instead of the sample of $n$ neurons that was actually recorded.
This is our \enquote{full network}.\mynote{how about calling it
  \enquote{the pool}?} The normalized total activity of these $N$ neurons
at $t$ is $\yA_{t}$. The relative frequencies of the various activity
levels during the recording were $(\yFF_{\yA}) \defs \yF$. We don't know
the values $\yA_{t}$ at each $t$, or the frequency distribution $\yF$. We
only know for certain that $\yA_{t} \in \set{0,1/N,\dotsc,(N-1)/N,1}$ and
that $N\yA_{t} \ge n\ya_{t}$ for obvious reasons. For the time being we
assume that we know $N$; in \sect*** we discuss the consequences of our
lack of precise knowledge about this number.

@@

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those
expectations to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First remark: the
distribution $\px(\yA)$ given by this method is the zeroth-order
approximation \citep[in the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four completely different quantities. Note that the four
distributions above numerically differ in higher-order approximations.
Second remark: the maximum-entropy method based on the Shannon entropy
implicitly makes some assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We say more about
these two points in \sect***.

In our case, to apply the maximum-entropy prescriptions to the total
activity of the full network we need to fix some averages of its belief
distribution, for example the distribution's moments. But we don't have any
measured moments for the full network to equate the distribution moments
to. Here enters point~\ref{item:sample_step}: the probability calculus
gives an exact, linear relation between the first $m$ moments for the full
network and the first $m$ for the sample
\citep[\eqns~(16)]{portamanaetal2015}; the ones determine the others and
vice versa at every time bin. This relation is a classical result of
sampling theory
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{n\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. In formulae these
constraints on $\px(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}}_{\text{measured moments}}
  =
   \underbracket{\sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)}_{\text{distribution moments}}
\end{equation}
for all $m$ we wish.

The calculation amounts to a convex optimization
\citep{meadetal1984}[\chap~10]{pressetal1988_r2007}{fangetal1997,boydetal2004_r2009,portamana2017b}
and for the numbers $N,n,m$ considered in the present work it can be done
on a modern computer without approximations of normalization constants or
potential functions.

@@

The number of moments used with this method depends on the questions and
hypotheses that a researcher is exploring; for example, hypotheses of
sufficient statistics, such as the sufficiency of pairwise correlations to
quantify our \dob\ about network activity. We discuss this kind of use in
\sect~\ref{sec:marginalization}.%, but our main goal is to simply introduce
% the general method without entering into biological questions of this
% kind.

The particular case in which $n$ moments are constrained is especially
important: it corresponds to fully constraining the marginal frequency
distribution for the activity of the sample neurons, $\yf$. In this case,
our belief about the full-network activity is based on all available
measured frequency data. Note that the application of the maximum-entropy
method \emph{at the sample level} is trivial and meaningless in this case
-- it just gives back the measured frequency distribution. But application
of the method \emph{at the level of the full network} is not trivial.

@@

In the next section we apply the method just described to the data sets
from two actual recordings, using six moments, and discuss the properties
of the resulting distributions.
